Ensemble Stability of Biological Networks

### 2014-09-24 paper.tex
update description of resampling multiple interaction weights

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices connectivity
$\connectivity_i$ and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, the probability that the system will remain stable after
resamling is $R(\hbox{avg stab} C_i, 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} G, 1) =
\frac{\sum_{i,j \in \hier(G)} \ell_{ij} +
      \sum_{i=1}^n \connectivity_i R(C_i)}
     {\sum_{i,j \in \hier(G)} \ell_{ij} +
      \sum_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat}.

Next, we derive an upper bound on the robustness of any graph constructed from a given set of connected components.  Since $R(C_i) \le 1$ for all $i$, it follws that increasing $\ell_{ij}$ will increase $R(G)$.  Given two connected components $C_i$ and $C_j$ with $v_i$ and $v_j$ nodes respectively, we have a maximum of $n_i n_j$ links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so, we conclude that
\begin{align}
R(G) \le \frac{\sum_{i,j \in \hier(G)} v_i v_j + \sum_i \connectivity_i R(C_i)}
{\sum_{i,j \in \hier(G)} v_i v_j + \sum_i \connectivity_i}, \label{eq:sccmaxrobustness}
\end{align}

Since every acyclic digraph can be embedded into a totally ordered set, we may assume without loss of generality that our components have been ordered in a way such that, if $(i,j) \in \hier(G)$, then $i < j$.  Hence, after substituting into our expression and simplifying using the identity
$$\sum_{i=1}^{n-1}\sum_{j=i+1}^{n}v_i v_j~=~\frac{1}{2} \left( \sum_{i=1}^{n} v_i \right)^2-\frac{1}{2} \sum_{i=1}^{n} v_i^2,$$
we arrive at the following upper bound on robustness in terms of the connected components:
\begin{align}
R_{\mathrm{max}}(C_1, \ldots C_n) =
\frac{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i \connectivity_i R(C_i)}
{\frac{1}{2} ((\sum_i v_i)^2 - \sum_i v_i^2) + \sum_i \connectivity_i}, \label{eq:exsccmaxrobustness}
\end{align}
Furthermore, this bound is attained.  Suppose that $G_{\mathrm{tot}}$ is the graph on $n$ nodes with a link from node $i$ to node $j$ whenever $i < j$.  Then, by the construction described in the previous section on network hierachy, we have a graph $G_{\mathrm{max}}$ such that the components of $G_{\mathrm{max}}$ are $C_1, \ldots C_n$ and $\hier (G_{\mathrm{max}}) = G_{\mathrm{tot}}$.  By our formula, $R(G_{\mathrm{max}}) = R_{\mathrm{max}}(C_1, \ldots C_n)$.

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that, as
before, we have a graph $G$ with strongly connected components $C_i$,
but that now we
resample $m$ nodes.  Then, given a partition $m = \sum_{i=0}^n m_i$
such that $m_0 \le \sum_{i=1}^{n-1}\sum_{j=i}^n \ell_{ij}$ and $m_i 
\le \connectivity_i$ when $1 \le i \le n$,
there are ${m \choose m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$
links from $C_i$ and $m_0$ links between strongly connected
components.  Hence, if we denote the set of partitions which satisfy
our inequalities as $M$, our weighted average becomes
\begin{equation}
R(\hbox{avg stab} (G), m) =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M} {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum_{i=1}^n m_i R(\hbox{avg stab} (C_i), m_i)\right)}
     {m \sum\limits_{(m_0, m_1, \ldots m_n) \in M} {m \choose m_0, m_1, \ldots. m_n}}
\end{equation}

The preceding argument demonstrates the manner in which robustness is maximized by maximizing the number of edges \emph{between} the strongly connected components of the graph underlying system interdependencies. And, the graph that has the latter property is that associated to the total ordering, $G_{\mathrm{tot}}$, which aslo possess the highest degree of hierarchy. The fact that robustness is equivalent for any configuration of strongly connected components with an equivalent number of links between them demonstrates that it is invariant to permutations of strongly connected components. This result does not depend at all upon any particular details such as the probability distribution from which the component interaction strengths are sampled because the arguments presented here are purely topological in nature. The result that dynamical robustness is correlated with system hierarchy therefore applies to an even broader class of dynamical systems than the particular random ensembles primarily studied here.

\todo{write down analogs of \ref{eq:sccrobustness} and \ref{eq:sccmaxrobustness} for resampling a fixed proportion or an arbitrary number of components}

>>>>

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices connectivity
$\connectivity_i$ and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $R(\hbox{avg stab} (C_i), 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} (G), 1) =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i R(C_i)}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

Next, we derive an upper bound on the robustness of any graph
constructed from a given set of connected components.  Since
$R(C_i) \le 1$ for all $i$, it follows that increasing $\ell_{ij}$ will
increase $R(G)$.  Given two connected components $C_i$ and $C_j$ with
$v_i$ and $v_j$ nodes respectively, we have a maximum of $n_i n_j$
links going from $C_i$ to $C_j$.  Hence, $\ell_{ij} \le v_i v_j$, so,
we conclude that
\begin{align}
R(G) \le \frac{\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i R(C_i)}
              {\sum\limits_{(i,j) \in \hier(G)} v_i v_j +
               \sum\limits_{i=1}^n \connectivity_i},
\end{align}

Since every acyclic digraph can be embedded into a totally ordered
set, we may assume without loss of generality that our components have
been ordered in a way such that, if $(i,j) \in \hier(G)$, then $i <
j$.  Hence, after substituting into our expression and simplifying
using the identity
$$\sum_{i=1}^{n-1}\sum_{j=i+1}^{n}v_i
v_j~=~\frac{1}{2} \left( \sum_{i=1}^{n} v_i \right)^2-\frac{1}{2} \sum_{i=1}^{n}
v_i^2,$$
we arrive at the following upper bound on robustness in terms
of the connected components:
\begin{align} \label{eq:sccmaxrobustness}
R_{\mathrm{max}}(C_1, \ldots C_n) =
\frac{\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i R(\hbox{avg stab}(C_i),1)}
     {\frac{1}{2} ((\sum_{i=1}^n v_i)^2 - \sum_{i=1}^n v_i^2) +
                    \sum_{i=1}^n \connectivity_i}.
\end{align}
Furthermore, this bound is attained.  Suppose that $G_{\mathrm{tot}}$
is the graph on $n$ nodes with a link from node $i$ to node $j$
whenever $i < j$.  Then, by the construction described in the previous
section on network hierachy, we have a graph $G_{\mathrm{max}}$ such
that the components of $G_{\mathrm{max}}$ are $C_1, \ldots C_n$ and
$\hier (G_{\mathrm{max}}) = G_{\mathrm{tot}}$.  By our formula,
$R(\hbox{avg stab} (G_{\mathrm{max}}), 1) = R_{\mathrm{max}}(C_1, \ldots C_n)$.

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
\forall i \in \{1, \ldots, n\} \, m_i < \connectivity_i \right\}.
\end{equation*}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{equation}
R(\hbox{avg stab} (G), m) =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i R(\hbox{avg stab} (C_i), m_i) \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}

As before, since $R(\hbox{avg stab} (C_i), m_i) \le 1$, we may
increase $R(\hbox{avg stab} (G), m)$ by increasing the maximum
possible value of $m_0$ while keeping the strongly connected
components the same.  Again, if we fix $\hier(G)$, the maximum
possible value of $m_0$ is $\sum_{(i,j) \in \hier(G)} v_i v_j$
whereas, if we allow it to vary, the maximum is $\frac{1}{2} ((\sum_{i=1}^n
v_i)^2 - \sum_{i=1}^n v_i^2)$, which is attained when $\hier(G_{max}) =
G_{tot}$.  Hence, we conclude that $R(\hbox{avg stab} (G), m) \le
R(\hbox{avg stab} (G_{\mathrm{max}}), m)$.

The preceding argument demonstrates the manner in which robustness is
maximized by maximizing the number of edges \emph{between} the strongly
connected components of the graph underlying system interdependencies.
And, the graph that has the latter property is that associated to the
ordering, $G_{\mathrm{tot}}$, which aslo possess the highest degree of
hierarchy. The fact that robustness is equivalent for any configuration
of strongly connected components with an equivalent number of links
between them demonstrates that it is invariant to permutations of
strongly connected components. This result does not depend at all upon
any particular details such as the probability distribution from which
the component interaction strengths are sampled because the arguments
presented here are purely topological in nature. The result that
dynamical robustness is correlated with system hierarchy therefore
applies to an even broader class of dynamical systems than the particular
random ensembles primarily studied here.
